---
title: " ITP-321 Final Project"
author: "Nacionales Maranello, Francis John Carlo, Ashlie Mae Orlanda"
date: "2026-05-17"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
```

#  Introduction

This project demonstrates core R programming concepts including loops, functions, data visualization, and introductory data science techniques. Each section includes working examples and explanations to highlight practical applications.

---

#  For Loop Scripts

## Staircase Pattern

This example demonstrates how a `for` loop can be used to build structured output step-by-step.

```{r}
build_staircase <- function(total_steps = 10L) {
  staircase <- ""
  for (step in seq_len(total_steps)) {
    staircase <- paste0(staircase, "Step ", step, ": ", paste(rep("*", step), collapse = " "), "\n")
  }
  staircase
}

cat(build_staircase(10))
```

---

## Multiplication Table Report

This script generates a formatted multiplication table and adds simple number insights.

```{r}
is_perfect_square <- function(x) {
  x >= 0 & floor(sqrt(x))^2 == x
}

for (i in 1:10) {
  product <- 7 * i
  parity <- ifelse(product %% 2 == 0, "EVEN", "ODD")
  note <- ifelse(is_perfect_square(product), " (square)", "")
  cat(i, "x 7 =", product, "-", parity, note, "\n")
}
```

---

#  While Loop Scripts

## Launch Simulation

Demonstrates how a `while` loop updates values dynamically over time.

```{r}
t <- 1
altitude <- 0
fuel <- 100

while (t <= 5) {
  altitude <- altitude + t * 10
  fuel <- fuel - 5
  cat("Time:", t, "| Altitude:", altitude, "| Fuel:", fuel, "\n")
  t <- t + 1
}
```

---

## Auction Countdown

A simple countdown using a condition-based loop.

```{r}
time <- 5
while (time > 0) {
  cat("Time left:", time, "\n")
  time <- time - 1
}
```

---

#  Functions

## Add Numbers

```{r}
add_numbers <- function(a, b) {
  a + b
}

add_numbers(5, 3)
```

---

## Rectangle Area

```{r}
calculate_area <- function(length, width) {
  length * width
}

calculate_area(5, 4)
```

---

#  Data Visualization

## Scatter Plot (mtcars)

This plot shows the relationship between car weight and fuel efficiency.

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "#2C3E50", size = 3) +
  theme_minimal() +
  labs(
    title = "Weight vs Fuel Efficiency",
    x = "Weight",
    y = "MPG"
  )
```

---

## Bar Chart (iris)

Displays the distribution of species in the dataset.

```{r}
ggplot(iris, aes(x = Species, fill = Species)) +
  geom_bar() +
  theme_minimal() +
  guides(fill = "none") +
  labs(title = "Count of Iris Species")
```

---

#  Data Science

## Linear Regression

This model predicts MPG based on weight.

```{r}
model <- lm(mpg ~ wt, data = mtcars)
summary(model)

ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal()
```

---

## Group Summary (iris)

Compares average petal length across species.

```{r}
iris_summary <- iris %>%
  group_by(Species) %>%
  summarise(avg_length = mean(Petal.Length))

iris_summary

ggplot(iris, aes(Species, Petal.Length, fill = Species)) +
  stat_summary(fun = mean, geom = "col") +
  theme_minimal() +
  guides(fill = "none")
```

---

# Conclusion

This project highlights foundational programming and data analysis skills in R, including control structures, reusable functions, data visualization, and basic statistical modeling. These techniques form the building blocks for more advanced data science workflows.